Nail: A practical interface generator for data formats

### 2014-09-11 paper/osdi14/design.tex
design

Nail grammars describe both the \emph{external format} and an
\emph{internal representation} of a protocol.
Nail produces the following from a single, descriptive grammar:

\begin{itemize}
\item \textit{Type declarations} for the internal model, which
the application should use to represent data items in memory.

\item The \textit{parser}, a function that the application
should invoke to parse a sequence of bytes into an instance
of the above model.

\item The \textit{generator}, a function that the application
should invoke to create a sequence of bytes from an instance
of the model.
\end{itemize}

For example, Figure~\ref{fig:dns-full} shows a Nail grammar for DNS
packets.  For this grammar, Nail produces the type declarations
shown in Figure~\ref{fig:dns-full-struct}, and the parser and
generator functions shown in Figure~\ref{fig:dns-full-api}.

@@

\paragraph{Semantic bijection.}
Parsing inputs and generating outputs suggests a bijection
between external data and its internal representation. However,
a bijection in the traditional sense often does not make sense for data
formats. Consider a grammar for a text language that tolerates white space, or a binary protocol
that tolerates arbitrarily long padding. Program semantics should be independent of the number of padding elements in the input, and Nail therefore
does not expose that information to the programmer. We call such discarded fields \emph{constants}. 
Similarly, programs should not necessarily preserve the layout of objects referred to by their offsets.

Therefore, Nail  establishes only a  \emph{semantic bijection} between the external format and the internal model.
That is, when Nail parses an input into an internal representation, and then
generates output from that representation, the two byte streams (input and
output) will have the same meaning (i.e., be interpreted equivalently by Nail).
However, the byte streams might not be identical. If the grammar
consists of a simple protocol without offset fields, constants, and the like, there is a
conventional bijection between  internal models and valid parser inputs.
%\footnote{Say, the physical layer of most communication
%  protocols is a possibly infinite sequence of symbols that are syntactically nil followed by a
%  pre-determined synchronization sequence and the actual contents of the transmission.}

\paragraph{Hiding redundant information.}
Nail's internal model is designed to hide unneeded and redundant information from the application.
Nail introduces \emph{dependent fields}, which contain data that can be computed during generation
and need to be kept as additional state during parsing. Dependent fields are, for example, used to represent
lengths, offsets, and checksums. If dependent fields were exposed in the internal model, information
would be duplicated and inconsistent internal data structures could be produced when data is
modified. For example, when using Nail to handle UDP packets, without dependent fields, programmers
might forget to update checksum fields when they modify the payload data.

\paragraph{Parser extensions.}
Real-world protocols contain complicated ways of encoding data. Fully representing these in an
intentionally limited model such as our parser language is impractical. Therefore, Nail introduces
\emph{transformations}, which allow arbitrary code by the programmer to interact with the parser and
generator. Nail parsers and generators interact with data through an abstract stream, which allows
reading and writing of bits and re-positioning. Transformations allow the programmer to write
functions in a general-purpose language that consume streams and define new temporary streams, while
also reading or writing the values of dependent fields.

@@

Initial versions of Nail's design included a special combinator for handling offset fields,
which consumed a dependent field and applied
a parser at the offset specified therein. However, it proved impossible to foresee all the ways in which a protocol could encode an offset;
for example, some protocols such as PDF and ZIP locate structures by scanning for a magic number starting at the
end of the file or at a fixed offset. In nested grammars, offsets are also not necessarily computed from
the beginning of a file or packet.
Nail's transformations allow the programmer to write arbitrary functions that can
handle such structures and streams, which are a generic abstraction for input and
output data that allow the decoded data to be integrated with the rest of the
generated Nail parser.

@@

\paragraph{Rules.}
A Nail grammar consists of rules that assign a parser to a name. Rules are written as assignments,
such as \cc{ints = /*parser definition*/}, which defines a rule called \cc{ints}. As we will
describe later in \S\ref{s:dependent} and \S\ref{s:transforms}, rules can optionally consume parameters.
Rules can be invoked in a Nail grammar anywhere a parser can appear. Rule invocations act as though
the body of the rule had been substituted in the code. If parameters appear, they are passed by
reference.
\paragraph{Integers and constraints.}

Nail's fundamental parsers represent signed or unsigned
integers with arbitrary lengths up to 64 bits.
Note that is possible to define parsers for sub-byte lengths, for example, the flag bits in the DNS
message header, in lines 5 through 8.

@@

The \cc{many} combinator takes a parser and applies it repeatedly
until it fails, returning an array of the inner parser's results. In line 39 of the DNS grammar, a
sequence of labels is parsed by parsing as many labels as possible, that is, until an invalid length
field is encountered.
The \cc{sepBy} combinator
additionally takes a constant parser, which it applies in between parsing
two values, but not before parsing the first value or after parsing the
last.  This is useful for parsing an array of items delimited by a separator.
%For examle, \cc{many uint8} represents an array of 8-bit unsigned
%integers, and \cc{sepBy uint8=',' (many uint8 | '0'..'9')} recognizes
%comma-separated lists of decimal numbers.

@@

\paragraph{Optional.}
Nail includes an \cc{optional} combinator, which attempts to recognize a value, but succeeds
without consuming input when it cannot recognize that value. Syntactically, \cc{optional} is
equivalent to a choice between the parser and an empty structure, but in the internal model it is
more concisely represented as a reference that is null when the parser fails.
For example, the grammar for  Ethernet headers uses \cc{optional vlan\_header} to parse the VLAN
header that appears only in Ethernet packets transmitted to a non-default VLAN\@.

\paragraph{References.}
Rules allow for recursive grammars. To support recursive data types, we introduce the
reference combinator \cc{*}  that does not change the syntax of the external format described, but
introduces a layer of indirection, such as a reference or pointer, to the model data type.
The reference combinator does not need to be used when another combinator, such as \cc{optional} or
\cc{many}, already introduces indirection in the data type. An example is shown in Figure~\ref{fig:grammar-arith}.

>>>>

Nail grammars describe both the \emph{external format} and an
\emph{internal representation} of a protocol, as opposed to existing
grammar languages, which describe only the former and leave it up to
the programmer to construct an internal representation, typically in
the form of an abstract syntax tree.
Nail produces the following from a single, descriptive grammar:

\begin{CompactItemize}
\item \textit{type declarations} for the internal model,
\item the \textit{parser}, a function to parse a sequence of bytes into an
instance of the model, and
\item the \textit{generator}, a function to create a
sequence of bytes from an instance of the model.
\end{CompactItemize}

For example, Figure~\ref{fig:dns-full} shows an example grammar for DNS packets. Nail produces two
functions shown in Figure~\ref{fig:dns-full-api} that parse a DNS packet into the structures
partially shown in
Figure~\ref{fig:dns-full-struct} and vice versa.

@@

\paragraph{Semantic bijection.}
Parsing inputs and generating outputs form a bijection. However,
a bijection in the traditional sense often does not make sense for data
formats. Consider a grammar for a text language that tolerates white space, or a binary protocol
that tolerates arbitrarily long padding. Program semantics should be independent of the number of padding elements in the input, and Nail therefore
does not expose that information to the programmer. We call such discarded fields \emph{constants}. 
Similarly, programs should not necessarily preserve the layout of objects referred to by their offsets.

Therefore, Nail  maintains only a  \emph{semantic bijection} between the external format and the internal model.
That is, when Nail parses an input into an internal representation, and then
generates output from that representation, the two byte streams (input and
output) will have the same meaning (i.e., be interpreted equivalently by Nail).
However, the byte streams might not be identical. If the grammar
consists of a simple protocol without offset fields, constants, and the like, there is a
conventional bijection between  internal models and valid parser inputs.
%\footnote{Say, the physical layer of most communication
%  protocols is a possibly infinite sequence of symbols that are syntactically nil followed by a
%  pre-determined synchronization sequence and the actual contents of the transmission.}

\paragraph{Hide redundant information.}
Nail's internal model is designed to hide unneeded and redundant information from the application.
Nail introduces \emph{dependent fields}, which contain data that can be computed during generation
and need to be kept as additional state during parsing. Dependent fields are, for example, used to represent
lengths, offsets, and checksums. If dependent fields were exposed in the internal model, information
would be duplicated and inconsistent internal data structures could be produced when data is
modified. For example when using Nail to handle UDP packets, without dependent fields, programmers
might forget to update checksum fields when they modify the payload data.

\paragraph{Parser extensions.}
Real-world protocols contain complicated ways of encoding data. Fully representing these in an
intentionally limited model such as our parser language is impractical. Therefore, Nail introduces
\emph{transformations}, which allow arbitrary code by the programmer to interact with the parser and
generator. Nail parsers and generators interact with data through an abstract stream, which allows
reading and writing of bits as re-positioning. Transformations allow the programmer to write
functions in a general purpose language that consume streams and define new temporary streams, while
also reading or writing the values of dependent fields.

Initial versions of Nail's design included a special combinator for handling offset fields,
which consumed a dependent field and applied
a parser at the offset specified therein. However, it proved impossible to foresee all the ways in which a protocol could encode an offset;
for example, some protocols such as PDF and ZIP locate structures by scanning for a magic number starting at the
end of the file or a fixed offset. In nested grammars, offsets are also not necessarily computed from
the beginning of a file or packet.
Nail's transformations allow the programmer to write arbitrary functions that can
handle such structures and streams, which are a generic abstraction for input and
output data that allow the decoded data to be integrated with the rest of the
generated Nail parser.

@@

@@

\paragraph{Parsing.}
\paragraph{Rules.}
A Nail grammar consists of rules that assign a parser to a name. Rules are written as assignments,
such as \cc{ints = /*parser definition*/}, which defines a rule called \cc{ints}. As we will
describe later in \S\ref{s:dependent} and \S\ref{s:transforms}, rules can optionally consume parameters.
Rules can be invoked in a Nail grammar anywhere a parser can appear. Rule invocations act as though
the body of the rule had been substituted in the code. If parameters appear, they are passed by
reference.
\paragraph{Integers and constraints.}

Nail's fundamental parsers represent signed or unsigned
integers with arbitrary lengths up to 64 bits.
Note that is possible to define parsers for sub-byte lengths, for example, the flag bits in the DNS
message header, in lines 4 through 8.

@@

The \emph{many} combinator takes a parser and applies it repeatedly
until it fails, returning an array of the inner parser's results. In line 39 of the DNS grammar, a
sequence of labels is parsed by parsing as many labels as possible, that is, until an invalid length
field is encountered.
The \emph{sepBy} combinator
additionally takes a constant parser, which it applies in between parsing
two values, but not before parsing the first value or after parsing the
last.  This is useful for parsing an array of items delimited by a separator.
%For examle, \cc{many uint8} represents an array of 8-bit unsigned
%integers, and \cc{sepBy uint8=',' (many uint8 | '0'..'9')} recognizes
%comma-separated lists of decimal numbers.

@@

\paragraph{Optional.}
Nail includes an \cc{optional} combinator, which attempts to recognize a value, but succeeds
without consuming input when it cannot recognize that value. Syntactically, \cc{optional} is
equivalent to a choice between the parser and an empty structure, but in the internal model it is
more concisely represented as a reference that is null when the parser fails.
For example, the grammar for  Ethernet headers uses \cc{optional vlan\_header} to parse the VLAN
header that only appears in Ethernet packets transmitted to a non-default VLAN\@.

\paragraph{References.}
Rules allow for recursive grammars. To support recursive data types, we allow introduce the
reference combinator \cc{*}  that does not change the syntax of the external format described, but
introduces a layer of indirection, such as a reference or pointer, to the model data type.
The reference combinator does not need to be used when another combinator, such as \cc{optional} or
\cc{many}, already introduces indirection in the data type. An example is shown in Figure~\ref{fig:grammar-arith}.
